Measuring contrast processing in the visual system using SSVEP

### 2024-05-10 review.qmd
No changes in text
Fixed a few minor typos

Because SSVEP responses at detection threshold are very small, estimating a threshold is achieved by measuring the contrast response function at relatively high levels, and extrapolaKting back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis (see @fig-sweepvep). This contrast level was shown to correspond approximately with psychophysically measured detection thresholds [@Norcia1986].

@@

A robust estimate of the threshold therefore requires the measurement of the SSVEP amplitude at many different super-threshold contrasr levels. This was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@Tyler1979]. To avoid hysteresis effects, the sweep is sometimes conducted both up and down in the same experiment [@Tyler1979, @Norcia1985a; @Norcia1990]. The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffmann2017; @Bach2008; @Bach2020].

@@

Recent work on understanding Parkinson's disease has also used *Drosophila* genetic models. @Afsari2014 found that mutant flies produced stronger SSVEP responses to flickering lights than control flies (see [@fig-plotclinical]d). The authors theorised that differences in early gain control during development might lead to visual deficits later in life. Although visual responses are a convenient assay of neural function, it is likely that same general process applies throughout the whole brain, including in the motor system where the core Parkinson's symptoms (tremor, rigidity, slow movement) manifest. The SSVEP differences were reduced by a kinase inhibitor that targets the dopamine system, demonstrating how model organisms can be used to test new pharmacological treatments. SSVEP responses also provide a potential method to diagnose Parkinson's before any symptoms manifest, and to monitor the effect of treatments.

@@

Attention exerts a profound influence on visual performance: for example, allowing people to attend covertly to a spatial location improves their performance on a target detection task significantly [@Posner1980; @Cameron2002; @Carrasco2000; @Bashinski1980; @Morrone2004; @Pestilli2009]. In principle, this enhancement may be driven both by modulation of the underlying signal or noise characteristics, or by additional decision-theoretic factors such as reduction in spatial uncertainty [@Gould2007; @Petrov2006]. Early experiments with non-human primates showed little evidence for attentionally-driven changes in neuronal spike rates [@Luck1997; @Marcus2002; @McAdams1999; @Mehta2000a; @Mehta2000b], but with the advent of spatially-resolved human brain imaging methods in the late 1990s it became apparent that spatial attention was linked to frank changes in both fMRI BOLD responses 
[@Tootell1998; @Brefczynski1999; @Buracas2007; @Gandhi1999; @Kastner1999; @Li2008; @Murray2008; @Silver2007; @Somers1999] and EEG signals [@Ding2006; @Muller2000; @Morgan1996; @Muller1998].

@@

One widespread recent application of the SSVEP technique is in the design of brain-computer interfaces, which seek to control some aspect of a computer using neural signals. The high SNR and precise frequency resolution of SSVEPs make them an ideal candidate for this approach. Typical studies may involve presenting an array of stimuli at different contrast flicker frequencies, and having the participant select one either by overt attention (i.e. shifting fixation to foveate the selected stimulus) or covert attention (i.e. deploying attention to one stimulus whilst keeping fixated) [@Middendorf2000]. Because SSVEP signals are highly sensitive to both visual field position [@Ales2010; @Di-Russo2007] and attentional state [@Morgan1996; @Muller1998; @Verghese2012; @Lauritzen2010], the response to the selected stimulus will typically increase relative to the others, allowing it to be identified by an on-line algorithm. Because more than one stimulus frequency is generally present, this modulation and the associated changes in gain control will affect the entire pattern of self- and intermodulation terms, allowing the choice to be decoded by a multivariate pattern classifier.

This approach is primarily useful in situations that require the BCI to distinguish from among a small set of possibilties: for example, in early work, visual stimuli representing 'Left' and 'Right' commands in a flight simulator were distinguished robustly [@Middendorf2000]. Although SSVEP-based BCI interfaces typically do use contrast flicker, work in this field has largely focused on optimising the stimuli or decoders to increase decoding performance (e.g. ... ) and reducing visual fatigue associated with the long-term presentation of arrays of high-contrast flicker [@Diez2024] rather than studying visual contrast processing \textit{per se}. We therefore note in passing that using SSVEPs to improve our understanding of early contrast processing may yield benefits to this related field.

# Future directions
The SSVEP is a powerful tool for studying contrast processing. It provides a high SNR readout of neuronal activity that is unambigously linked to the input. It is sensitive to both the amplitude and phase of the input and when combined with source imaging, it can be extracted from different cortical regions allowing researchers to track contrast processing computations across the visual pathway. Because of the high SNR, it can be measured in subjects where long recording durations are impractical (for example, infants or patients with neurological disorders) and at frequencies high enough to be effectively invisible [@Herrmann2001; @Minarik2023; @Seijdel2023]. The SSVEP is also able to 'fingerprint' modulators of the inputs through the harmonic and intermodulation terms they generate in the output. In principle, each nonlinearity in the visual pathway can be identified by its contribution to the frequency spectrum at different recording locations [@Regan1988]. This, it turn, allows researchers to study how contrast processing depends on spatial and temporal context and as well as changes in task, cognitive and behavioural state and arousal.

@@

Finally, SSVEPs continue to provide a way of studying contrast in the cortex. Here, we are often interested in how different visual parameters interact. For example, how are signals from different eyes combined to generate both scalar contrast values and also binocular depth cues? How does contrast combination depend on the low level properties of the individual inputs such as their retinotopic location, cone contrast, spatiotemporal frequency, eye of origin and orientation? Colour vision scientists are particularly interested in how chromatic signals originating in a small number of cone-opponent retinal pathways are transformed into a perceptual colour space where the 'unique hues' appear to be only weakly-related to the early retinal outputs and how these computations are conditioned by the spatial and temporal properties of the scene [@Li2022; @Stoughton2008; @Kaneko2020; @Gegenfurtner2003; @Solomon2007; @Wandell1993]. All of these questions can be addressed by using SSVEP and frequency tagging to examine the computations that combine and transform the inputs at different cortical stages [@Watts2024; @Baker2017; @Chen2021; @Katyal2018; @Busse2009]. The original promise of the SSVEP approach was that the  entire complex-valued frequency spectrum recorded at each lcoation provided detailed information about the processing nonlinearities up to and including that point. In principle, this would allow researchers to uniquely fit the parameters of their computations models of visual processing [@Regan1988].  To date, the complexity of the neuronal computations at even the earliest stage of visual processing have hampered this effort - researchers typically restrict their analyses to the amplitudes of single, low-order frequency components (for example, the inputs frequencies or simple sums and differences of those frequencies). As our understanding of the early visual system improves, it is becoming possible to generate more realistic parameterised forward models of signal generation [@Chariker2020; @Groen2022; @Baker2017; @Tsai2012; @Schrimpf2020]. Feeding frequency-tagged inputs into these types of model allows us to generate synthetic SSVEP responses that can be compared with those measured in human subject. In principle, we are therefore able to use the SSVEP to fit the parameters of early visual processing. This approach may also allow us to develop more sensitive tests for the changes in early visual processing that accompany a wide range of neurological diseases and disorders.

>>>>

Because SSVEP responses at detection threshold are very small, estimating a threshold is achieved by measuring the contrast response function at relatively high levels, and extrapolating back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis (see @fig-sweepvep). This contrast level was shown to correspond approximately with psychophysically measured detection thresholds [@Norcia1986].

@@

A robust estimate of the threshold therefore requires the measurement of the SSVEP amplitude at many different super-threshold contrast levels. This was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@Tyler1979]. To avoid hysteresis effects, the sweep is sometimes conducted both up and down in the same experiment [@Tyler1979; @Norcia1985a; @Norcia1990]. The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffmann2017; @Bach2008; @Bach2020].

@@

Recent work on understanding Parkinson's disease has also used *Drosophila* genetic models. @Afsari2014 found that mutant flies produced stronger SSVEP responses to flickering lights than control flies (see [@fig-plotclinical]d). The authors theorised that differences in early gain control during development might lead to visual deficits later in life. Although visual responses are a convenient assay of neural function, it is likely that the same general process applies throughout the whole brain, including in the motor system where the core Parkinson's symptoms (tremor, rigidity, slow movement) manifest. The SSVEP differences were reduced by a kinase inhibitor that targets the dopamine system, demonstrating how model organisms can be used to test new pharmacological treatments. SSVEP responses also provide a potential method to diagnose Parkinson's before any symptoms manifest, and to monitor the effect of treatments.

@@

Attention exerts a profound influence on visual performance: for example, instructing people to attend covertly to a spatial location improves their performance on a target detection task significantly [@Posner1980; @Cameron2002; @Carrasco2000; @Bashinski1980; @Morrone2004; @Pestilli2009]. In principle, this enhancement may be driven both by modulation of the underlying signal or noise characteristics, or by additional decision-theoretic factors such as reduction in spatial uncertainty [@Gould2007; @Petrov2006]. Early experiments with non-human primates showed little evidence for attentionally-driven changes in neuronal spike rates [@Luck1997; @Marcus2002; @McAdams1999; @Mehta2000a; @Mehta2000b], but with the advent of spatially-resolved human brain imaging methods in the late 1990s it became apparent that spatial attention was linked to frank changes in both fMRI BOLD responses 
[@Tootell1998; @Brefczynski1999; @Buracas2007; @Gandhi1999; @Kastner1999; @Li2008; @Murray2008; @Silver2007; @Somers1999] and EEG signals [@Ding2006; @Muller2000; @Morgan1996; @Muller1998].

@@

One widespread recent application of the SSVEP technique is in the design of brain-computer interfaces, which seek to control some aspect of a computer using neural signals. The high SNR and precise frequency resolution of SSVEPs make them an ideal candidate for this approach. Typical studies may involve presenting an array of stimuli at different flicker frequencies, and having the participant select one either by overt attention (i.e. shifting fixation to foveate the selected stimulus) or covert attention (i.e. deploying attention to one stimulus whilst keeping fixated) [@Middendorf2000]. Because SSVEP signals are highly sensitive to both visual field position [@Ales2010; @Di-Russo2007] and attentional state [@Morgan1996; @Muller1998; @Verghese2012; @Lauritzen2010], the response to the selected stimulus will typically increase relative to the others, allowing it to be identified by an on-line algorithm. Because more than one stimulus frequency is generally present, this modulation and the associated changes in gain control will affect the entire pattern of self- and intermodulation terms, allowing the choice to be decoded by a multivariate pattern classifier.

This approach is primarily useful in situations that require the BCI to distinguish from among a small set of possibilties: for example, in early work, visual stimuli representing 'Left' and 'Right' commands in a flight simulator were distinguished robustly [@Middendorf2000]. Although SSVEP-based BCI interfaces typically do use contrast flicker, work in this field has largely focused on optimising the stimuli or decoders to increase decoding performance and reducing visual fatigue associated with the long-term presentation of arrays of high-contrast flicker [@Diez2024], rather than studying visual contrast processing \textit{per se}. We therefore note in passing that using SSVEPs to improve our understanding of early contrast processing may yield benefits to this related field.

# Future directions

The SSVEP is a powerful tool for studying contrast processing. It provides a high SNR readout of neuronal activity that is unambigously linked to the input. It is sensitive to both the amplitude and phase of the input and when combined with source imaging, it can be extracted from different cortical regions allowing researchers to track contrast processing computations across the visual pathway. Because of the high SNR, it can be measured in subjects where long recording durations are impractical (for example, infants or patients with neurological disorders) and at frequencies high enough to be effectively invisible [@Herrmann2001; @Minarik2023; @Seijdel2023]. The SSVEP is also able to 'fingerprint' modulators of the inputs through the harmonic and intermodulation terms they generate in the output. In principle, each nonlinearity in the visual pathway can be identified by its contribution to the frequency spectrum at different recording locations [@Regan1988]. This, it turn, allows researchers to study how contrast processing depends on spatial and temporal context, as well as changes in task, cognitive and behavioural state, and arousal.

@@

Finally, SSVEPs continue to provide a way of studying contrast in the cortex. Here, we are often interested in how different visual parameters interact. For example, how are signals from different eyes combined to generate both scalar contrast values and also binocular depth cues? How does contrast combination depend on the low level properties of the individual inputs such as their retinotopic location, cone contrast, spatiotemporal frequency, eye of origin and orientation? Colour vision scientists are particularly interested in how chromatic signals originating in a small number of cone-opponent retinal pathways are transformed into a perceptual colour space where the 'unique hues' appear to be only weakly-related to the early retinal outputs and how these computations are conditioned by the spatial and temporal properties of the scene [@Li2022; @Stoughton2008; @Kaneko2020; @Gegenfurtner2003; @Solomon2007; @Wandell1993]. All of these questions can be addressed by using SSVEP and frequency tagging to examine the computations that combine and transform the inputs at different cortical stages [@Watts2024; @Baker2017; @Chen2021; @Katyal2018; @Busse2009]. The original promise of the SSVEP approach was that the  entire complex-valued frequency spectrum recorded at each location provided detailed information about the processing nonlinearities up to and including that point. In principle, this would allow researchers to uniquely fit the parameters of their computational models of visual processing [@Regan1988].  To date, the complexity of the neuronal computations at even the earliest stage of visual processing have hampered this effort - researchers typically restrict their analyses to the amplitudes of single, low-order frequency components (for example, the input frequencies or simple sums and differences of those frequencies). As our understanding of the early visual system improves, it is becoming possible to generate more realistic parameterised forward models of signal generation [@Chariker2020; @Groen2022; @Baker2017; @Tsai2012; @Schrimpf2020]. Feeding frequency-tagged inputs into these types of model allows us to generate synthetic SSVEP responses that can be compared with those measured in human subjects. In principle, we are therefore able to use the SSVEP to fit the parameters of early visual processing. This approach may also allow us to develop more sensitive tests for the changes in early visual processing that accompany a wide range of neurological diseases and disorders.
